---
title: "Results"
editor: visual
---

## Tables! Lots and Lots of Tables!!

If you took the time to read through everything else on this site before coming here, then that was really cool of you and I appreciate you looking through my work! If you jumped straight to this tab, that's cool too! This is where all of the interesting stuff is anyways. I couldn't figure out how to add filters or whatever so you could create custom tables of selected years and statistics so this page presents tables with comparisons for all years combined, while single year tables can be found by hovering over the Results tab at the top of the page.

## Record

So how did the O's do? Not great! But if you follow baseball at all then you didn't need me to do all this work just to tell you that. Table 1 shows that overall the team won two-thirds of it's games during the non-breeding/wintering season, by far the best results of any of the breeding seasons, but only played six total games over ten years during that breeding season. They were a sub-.500 team in each of the other three breeding seasons, with the best winning percentage being in games played during the breeding season (\~48% of 476 games). The team had roughly the same winning percentage in the post-breeding migratory (\~45% of 565 games) and pre-breeding migratory (\~45% of 471 games). The obvious and undisputed conclusion here is that the breeding season brings outs the best in ALL of Baltimore's orioles.

```{r echo = FALSE, message = FALSE, warning = FALSE}

library(gt)
library(dplyr)
library(here)

orioles_log_final_all_years <-
  read.csv(here("data", "baseball", "orioles_log_final_all_years.csv"))

orioles_log_final_all_years |>
  group_by(bird_season, game_result) |>
  summarise(win_lose = n()) |>
  
  gt() |>
  tab_header(
    title = md("**Table 1. Baltimore Orioles Record by Breeding Season**") # md("** **") makes label bold
  ) |>
  cols_move_to_start(
    columns = c(win_lose)) |>
  cols_label(
    win_lose = "",
    game_result = ""
  )
```

Table 2 shows that the team played way more games within the breeding/migratory ranges of it's colorful winged talisman than outside of them, though the results were not particularly good either way: the O's won \~46% of 1,312 games in range and \~43% of 206 games out of range. Figure 1 presents a faceted map of each of the seasonal ranges for reference. Each of the dots represents an oriole sighting in the eBird data, with the borders of the ranges corresponding to the outermost observations. Only observations for 2012 are shown here but they were similarly distributed for each of the study years.

```{r echo = FALSE, message = FALSE, warning = FALSE}

library(gt)
library(dplyr)
library(here)

orioles_log_final_all_years <-
  read.csv(here("data", "baseball", "orioles_log_final_all_years.csv"))

orioles_log_final_all_years |>
  group_by(range_flag, game_result) |>
  summarise(win_lose = n()) |>
  
  gt() |>
  tab_header(
    title = md("**Table 2. Baltimore Orioles Record by Whether Game was Played In or Out of Seasonal Breeding Range**"), # md("** **") makes label bold
    subtitle = "2012 to 2021 MLB Seasons"
  ) |>
  cols_move_to_start(
    columns = c(win_lose)) |>
  cols_label(
    win_lose = "",
    game_result = ""
  )

```

```{r echo = FALSE, message = FALSE, warning = FALSE}

library(ggplot2)
library(rnaturalearth)
library(dplyr)
library(here)

birds_2012_sf <-
  read.csv(here("data", "birds", "birds_2012.csv"))

us_canada_fig1 <-
  ne_countries(scale = "medium", returnclass = "sf") |>
  filter(type == "Country" | type == "Sovereign country") |>
  filter(sovereignt == "United States of America" | sovereignt == "Canada")

birds_2012_sf_fig1 <-
  birds_2012_sf |>
  filter(state_perc > 0.0001)

ggplot() +
  geom_sf(data = us_canada_fig1) +
  coord_sf(xlim = c(-125.0, -55.0), ylim = c(20.0, 55.0)) +
  geom_point(data = birds_2012_sf_fig1, 
             aes(x = LONGITUDE, y = LATITUDE, color = "red")) +
  facet_wrap(~ season) +
  theme(legend.position = "none",
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks = element_blank()) +
  ggtitle(label = "Figure 1. Breeding and Migratory Ranges \nof the Baltimore Oriole, 2012") +
  xlab("") +
  ylab("")
```

Table 3 presents the Orioles' record stratified by both breeding season and whether the game was played in or out of the range. A bunch more losing records on display here, though they did have a 22-20 record when playing outside of the bird's pre-breeding migratory range during that season.

```{r echo = FALSE, message = FALSE, warning = FALSE}
library(gt)
library(dplyr)
library(here)

orioles_log_final_all_years <-
  read.csv(here("data", "baseball", "orioles_log_final_all_years.csv"))

orioles_log_final_all_years |>
  group_by(bird_season, range_flag, game_result) |>
  summarise(win_lose = n()) |>
  
  gt() |>
  tab_header(
    title = md("**Table 3. Baltimore Orioles Record by Breeding Season and Whether Game Was Played In or Out of the Seasonal Range**"), # md("** **") makes label bold
    subtitle = "2012 to 2021 MLB Seasons"
  ) |>
  cols_move_to_start(
    columns = c(win_lose)) |>
  cols_label(
    win_lose = "",
    game_result = ""
  )
```

## Batting Stats

If you're looking at this site then you've clearly been dying to know for years, decades even, whether them O's are also swinging the hot wood during the mascot's breeding season (I'm not sorry). Well look no further! There's too much to cover in these tables, and I didn't run any statistical tests to highlight potentially interesting comparisons (if Gunnar Henderson doesn't need t-tests then neither do I!), so just browse all of the tables below at your leisure.

```{r echo = FALSE, message = FALSE, warning = FALSE}

library(gt)
library(dplyr)
library(here)

orioles_log_final_all_years <-
  read.csv(here("data", "baseball", "orioles_log_final_all_years.csv"))

orioles_log_final_all_years |>
  group_by(bird_season, range_flag) |>
  summarise(games = n(),
            mean_bal_BA = mean(bal_BA),
            mean_opp_BA = mean(opp_BA),
            mean_bal_h = mean(bal_H),
            mean_opp_h = mean(opp_H),
            mean_bal_doub = mean(bal_doubles),
            mean_opp_doub = mean(opp_doubles),
            mean_bal_trip = mean(bal_triples),
            mean_opp_trip = mean(opp_triples),
            mean_bal_hr = mean(bal_HR),
            mean_opp_hr = mean(opp_HR)
            ) |>
  gt() |>
  tab_header(
    title = md("**Table 4a. Comparison of Selected Batting Stats Stratified by Breeding Season and Breeding/Migratory Range**"), # md("** **") makes label bold
    subtitle = "2012 to 2021 MLB Seasons"
  ) |>
  cols_move_to_start(
    columns = c(games)
  ) |>
  cols_label(
    range_flag = html("Season & In <br>or Out of Range"),
    games = html("# of <br>Games"),
    mean_bal_BA = html("Mean <br>BAL BA"),
    mean_opp_BA = html("Mean <br>Opp. BA"),
    mean_bal_h = html("Mean <br>BAL Hits"),
    mean_opp_h = html("Mean <br>Opp. Hits"),
    mean_bal_doub = html("Mean <br>BAL Doubles"),
    mean_opp_doub = html("Mean <br>Opp. Doubles"),
    mean_bal_trip = html("Mean <br>BAL Triples"),
    mean_opp_trip = html("Mean <br>Opp. Triples"),
    mean_bal_hr = html("Mean <br>BAL HR"),
    mean_opp_hr = html("Mean <br>Opp. HR"),
  ) |>
  fmt_number(
    columns = c(mean_bal_BA, mean_opp_BA),
    decimals = 3
  ) |>
  fmt_number(
    columns = c(mean_bal_h, mean_opp_h,
                mean_bal_doub, mean_opp_doub,
                mean_bal_trip, mean_opp_trip,
                mean_bal_hr, mean_opp_hr),
    decimals = 2
  ) |>
  fmt_number(
    columns = games,
    decimals  = 0
  )
```

```{r echo = FALSE, message = FALSE, warning = FALSE}

library(gt)
library(dplyr)
library(here)

orioles_log_final_all_years <-
  read.csv(here("data", "baseball", "orioles_log_final_all_years.csv"))

orioles_log_final_all_years |>
  group_by(bird_season, range_flag) |>
  summarise(games = n(),
            mean_bal_runs = mean(bal_runs),
            mean_opp_runs = mean(opp_runs),
            mean_bal_bb = mean(bal_BB),
            mean_opp_bb = mean(opp_BB),
            mean_bal_k = mean(bal_hitters_K),
            mean_opp_k = mean(opp_hitters_K),
            mean_bal_sb = mean(bal_SB),
            mean_opp_sb = mean(opp_SB),
            mean_bal_lob = mean(bal_LOB),
            mean_opp_lob = mean(opp_LOB)
            ) |>
  gt() |>
  tab_header(
    title = md("**Table 4b. Comparison of Selected Batting Stats Stratified by Breeding Season and Breeding/Migratory Range**"), # md("** **") makes label bold
    subtitle = "2012 to 2021 MLB Seasons"
  ) |>
  cols_move_to_start(
    columns = c(games)
  ) |>
  cols_label(
    range_flag = html("Season & In <br>or Out of Range"),
    games = html("# of <br>Games"),
    mean_bal_runs = html("Mean <br>BAL Runs"),
    mean_opp_runs = html("Mean <br>Opp. Runs"),
    mean_bal_bb = html("Mean <br>BAL Walks"),
    mean_opp_bb = html("Mean <br>Opp. Walks"),
    mean_bal_k = html("Mean <br>BAL Ks"),
    mean_opp_k = html("Mean <br>Opp. Ks"),
    mean_bal_sb = html("Mean <br>BAL SBs"),
    mean_opp_sb = html("Mean <br>Opp. SBs"),
    mean_bal_lob = html("Mean <br>BAL LOB"),
    mean_opp_lob = html("Mean <br>Opp. LOB")
  ) |>
  fmt_number(
    columns = c(mean_bal_runs, mean_opp_runs,
                mean_bal_bb, mean_opp_bb,
                mean_bal_k, mean_opp_k,
                mean_bal_sb, mean_opp_sb,
                mean_bal_lob, mean_opp_lob),
    decimals = 2
  ) |>
  fmt_number(
    columns = games,
    decimals  = 0
  )
```

## Pitching Stats

```{r echo = FALSE, message = FALSE, warning = FALSE}

library(gt)
library(dplyr)
library(here)

orioles_log_final_all_years <-
  read.csv(here("data", "baseball", "orioles_log_final_all_years.csv"))

orioles_log_final_all_years |>
  group_by(bird_season, range_flag) |>
  summarise(games = n(),
            mean_bal_er = mean(bal_team_ER),
            mean_opp_er = mean(opp_team_ER),
            mean_bal_wp = mean(bal_WP),
            mean_opp_wp = mean(opp_WP),
            mean_bal_balks = mean(bal_balks),
            mean_opp_balks = mean(opp_balks),
            mean_bal_n_pitchers = mean(bal_num_pitchers),
            mean_opp_n_pitchers = mean(opp_num_pitchers)
            ) |>
  gt() |>
  tab_header(
    title = md("**Table 5. Comparison of Selected Pitching Stats Stratified by Breeding Season and Breeding/Migratory Range**"), # md("** **") makes label bold
    subtitle = "2012 to 2021 MLB Seasons"
  ) |>
  cols_move_to_start(
    columns = c(games)
  ) |>
  cols_label(
    range_flag = html("Season & In <br>or Out of Range"),
    games = html("# of <br>Games"),
    mean_bal_er = html("Mean <br>BAL Earned Runs"),
    mean_opp_er = html("Mean <br>Opp. Earned Runs"),
    mean_bal_wp = html("Mean <br>BAL Wild Pitches"),
    mean_opp_wp = html("Mean <br>Opp. Wild Pitches"),
    mean_bal_balks = html("Mean <br>BAL Balks"),
    mean_opp_balks = html("Mean <br>Opp. Balks"),
    mean_bal_n_pitchers = html("Mean <br>BAL # Pitchers"),
    mean_opp_n_pitchers = html("Mean <br>Opp. # Pitchers")
  ) |>
  fmt_number(
    columns = c(mean_bal_er, mean_opp_er,
                mean_bal_wp, mean_opp_wp,
                mean_bal_balks, mean_opp_balks,
                mean_bal_n_pitchers, mean_opp_n_pitchers
                ),
    decimals = 2
  ) |>
  fmt_number(
    columns = games,
    decimals  = 0
  )

```

## Fielding Stats

```{r echo = FALSE, message = FALSE, warning = FALSE}

library(gt)
library(dplyr)
library(here)

orioles_log_final_all_years <-
  read.csv(here("data", "baseball", "orioles_log_final_all_years.csv"))

orioles_log_final_all_years |>
  group_by(bird_season, range_flag) |>
  summarise(games = n(),
            mean_bal_assists = mean(bal_assists),
            mean_opp_assists = mean(opp_assists),
            mean_bal_errors = mean(bal_errors),
            mean_opp_errors = mean(opp_errors),
            mean_bal_pass_balls = mean(bal_pass_balls),
            mean_opp_pass_balls = mean(opp_pass_balls),
            mean_bal_doub_plays = mean(bal_double_play),
            mean_opp_doub_plays = mean(opp_douple_play) #typo in var name
            ) |>
  gt() |>
  tab_header(
    title = md("**Table 6. Comparison of Selected Fielding Stats Stratified by Breeding Season and Breeding/Migratory Range**"), # md("** **") makes label bold
    subtitle = "2012 to 2021 MLB Seasons"
  ) |>
  cols_move_to_start(
    columns = c(games)
  ) |>
  cols_label(
    range_flag = html("Season & In <br>or Out of Range"),
    games = html("# of <br>Games"),
    mean_bal_assists = html("Mean <br>BAL Assists"),
    mean_opp_assists = html("Mean <br>Opp. Assists"),
    mean_bal_errors = html("Mean <br>BAL Errors"),
    mean_opp_errors = html("Mean <br>Opp. Errors"),
    mean_bal_pass_balls = html("Mean <br>BAL Pass Balls"),
    mean_opp_pass_balls = html("Mean <br>Opp. Pass Balls"),
    mean_bal_doub_plays = html("Mean <br>BAL Double Plays Turned"),
    mean_opp_doub_plays = html("Mean <br>Opp. Double Plays Turned")
  ) |>
  fmt_number(
    columns = c(mean_bal_assists, mean_opp_assists,
                mean_bal_errors, mean_opp_errors,
                mean_bal_pass_balls, mean_opp_pass_balls,
                mean_bal_doub_plays, mean_opp_doub_plays
                ),
    decimals = 2
  ) |>
  fmt_number(
    columns = games,
    decimals  = 0
  )


```
